Optimal updating magnitude in adaptive flat-distribution-sampling simulations

### 2017-08-19 doc/invt.tex
invt.tex: editing appendix A

%To simplify the ensuing discussion,
Second, we limit ourselves to matrices % $\mathbf w$'s
that satisfy
the detailed balance condition for $\pmb\rho$,
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}
  .
  \label{eq:w_detailedbalance}
\end{equation}
%
This requires the scaled updating matrix,
$\sqrt{ \rho_i/\rho_j } \, w_{ij}$,
to be symmetric,
and the symmetry allows the diagonalization
of $\mathbf w$ with a set of
eigenvectors, $\phi_{ki}$,
%
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij}
  =
  \lambda_k \, \phi_{kj}
  ,
\label{eq:eig_w}
\end{equation}
%
satisfying the orthonormal conditions\cite{vankampen}:
%
\begin{align}
  \sum_{k = 0}^{n - 1}
    \phi_{ki} \, \phi_{kj}
  &=
  \delta_{ij} \, \rho_i,
  \label{eq:eig_orthonormal_cols}
  \\
  \sum_{i = 1}^n
    \frac{ \phi_{ki} \, \phi_{li} }
         { \rho_i }
  &=
  \delta_{kl}
  ,
  \label{eq:eig_orthonormal_rows}
\end{align}
%
where we have started the index of the eigenmodes at $0$
instead of $1$ for later convenience.

@@

\item \label{step:alpha}
Compute the optimal schedule by
numerically integrating Eqs. \eqref{eq:q_opt} and \eqref{eq:mint}.
%
To handle Gaussian updating schemes more effectively,
we have chosen the $\mathcal N+1$ grid points
($\mathcal N=20000$ here)
over $[0, q(T)]$ to be given by
$q_{[i]} = \bigl(q(T) + 1\bigr) - \bigl(q(T)+1\bigr)^{1-i/\mathcal N}$,
where $i = 0, \dots, \mathcal N$.\footnote{For
  a mass distribution, $m(\bar q)$, that is
  inversely proportional to $\bar q$,
  this grid would make the integral,
  or the difference, $t_{[i+1]} - t_{[i]}$,
  roughly a constant.}
%
The curve $q(t)$ is then integrated
using the trapezoidal rule\cite{press3rd},
resulting in a list of values, $\bigl(q_{[i]}, t_{[i]} \bigr)$.
%
We then compute the updating magnitude by numerical differentiation:
$\alpha_i = (1 - e^{-\lambda_1 \, \hat \alpha_i})/\lambda_1$\footnote{This
  discretization follows from the runtime average interpretation
  discussed in Appendix \ref{sec:equilerr}.}
and
$\hat \alpha_i \equiv (q_{[i+1]} - q_{[i-1]}) / (t_{[i+1]} - t_{[i-1]})$,
with
$q_{[ -1]} \equiv 0$,
$q_{[\mathcal N+1]} \equiv q(T)$,
$t_{[ -1]} \equiv 0$, and
$t_{[\mathcal N+1]} \equiv T$.
%
The resulting values of $(t_{[i]}, \alpha_{[i]})$
can be linearly interpolated
to get the updating magnitude, $\alpha(t)$, at any simulation time, $t$.

@@

In addition to the above application to
homogeneous updating schemes,
we will consider here bandpass updating schemes
constructed from a sequence of orthogonal polynomials.
%
These updating schemes are more intuitive and easier to implement
as the bias potential in this case is
a low-order polynomial of the collective variable.

@@

For $k > 0$, $R_k(z)$ is a polynomial of degree $k$ with zero expectation.
%
Thus, $R_k\bigl(z(t)\bigr)$ roughly
represents the deviation of the $k$th statistical moment
of the instantaneous distribution
from its expectation.
Since high-order moments are generally less reliable,
this method is best used with a small $K$ for a narrow distribution.
%
However, by constructing a mixture\cite{shirts2017}
or expanded ensemble\cite{swendsen1986,
  *geyer1991, *hukushima1996, *hansmann1997, *sugita1999,
  *earl2005, *zuckerman2011, *rauscher2009,
  neuhaus2006, *neuhaus2007, kim2010,
  marinari1992,
  *lyubartsev1992, li2007,
  park2007, *nguyen2013, *zhang2015st},
we can simultaneously sample multiple target distributions
spread over a wider range, as exemplified below.

@@

In real application, the algorithm is often applied simultaneously
to an array of adjacent intervals
to cover a wider range of the collective variable, $z$.
%
In this way, an attempt to move out of the current interval
does not have to be rejected,
but can be interpreted as a transition to
another internal and accepted with a proper probability.

@@

\item
  Change the updating magnitude,
  which will be used for both intra-umbrella
  parameters, $c_k^{(i)}$,
  and inter-umbrella parameters, $\ln Z^{(i)}$.
  %

@@

After simulation,
the weighted histogram analysis method
(WHAM)\cite{ferrenberg1988, *ferrenberg1989,
*kumar1992, *roux1995, *bartels1997, *souaille2001,
*kobrak2003, *gallicchio2005, *chodera2007,
*fenwick2008, *kim2011, *shirts2008,
zhu2012, newman, frenkel},
can be used to recover the unbiased distribution.
%

@@

\section{\label{sec:equilerr}
Runtime average interpretation
of the inverse-time schedule
}

The optimality of the inverse-time schedule
for the single-bin updating scheme
can be understood from the observation that
the schedule allows the bias potential
to coincide with a runtime average
of independent corrections.

Assuming that we can linearize the histogram correction
in Eq. \eqref{eq:vcorr_equil} as
$\ln x \approx x - 1$ % (for $x \approx 1$),
and apply it to the instantaneous histogram, $h_i(t)$,
the independent estimate of bias potential in each step
would be given by
%
\begin{equation}
  \hat u_i(t) \approx u_i(t) + \frac{ h_i(t) } { \rho_i } - 1
  = u_i(t) + f_i(t) - 1
  .
  \label{eq:uhatt}
\end{equation}
%
Below we will use capital letters to denote
time averages, e.g. for variable $x$, we define
$X(t) \equiv \frac{1}{t} \sum_{\tau=1}^t x(\tau)$.
%\begin{equation}
%  X(t) \equiv \frac{1}{t} \sum_{\tau=1}^t x(\tau)
%  .
%  \notag
%  %\label{eq:timeav}
%\end{equation}
%
Time averaging Eq. \eqref{eq:uhatt} yields
\begin{align}
  \hat U_i(t)
  &\approx
  U_i (t)
  + F_i(t) - 1
  \approx
  U_i(t)
  + \ln \frac{ H_i(t) } { \rho_i }
  .
  \label{eq:uhatav}
\end{align}
%
%Equation \eqref{eq:uhatav}
This equation serves as
a generalization of Eq. \eqref{eq:vcorr_equil}
for FDS simulations under a variable bias potential
and we can use it to improve the estimated PMF
after simulation.\footnote{Equation \eqref{eq:uhatav}
  can also be derived
  from the self-healing umbrella sampling method\cite{marsili2006}.
  From Eq. \eqref{eq:uhatt},
  we can construct an independent estimator
  of the unbiased distribution as
  $\hat p_i(t) \approx h_i(t) \, e^{ u_i(t) }$.
  Upon time averaging, we get
  $\rho_i e^{\hat U_i(t)} = \hat P_i(t) \approx H_i \, e^{ U_i(t) }$.}
%
In particular, for the single-bin updating scheme,
Eq. \eqref{eq:wl_update},
we have, up to a constant shift,
%
\begin{align}
  \hat U_i(t)
  &\approx
  \frac{1}{t} \sum_{\tau=1}^t
  \left[
    u_i(\tau) +
    \frac{ u_i(\tau + 1) - u_i(\tau) } { \alpha(t) }
  \right]
  ,
  \notag
  \\
  &\approx
  \Delta u_i
  +
  \sum_{\tau=1}^t
  \left[
    1 + \frac{1}{\alpha(\tau - 1)} - \frac{1}{\alpha(\tau)}
  \right]
  \frac{ u_i(\tau) } {t}
  ,
  \label{eq:Uhat_sum}
\end{align}
where
$\Delta u_i \equiv u_i(t+1) /[t \, \alpha(t)]
- u_i(1)/[t \, \alpha(0)]$.
%$$
%\Delta u_i \equiv \frac{u_i(t+1)}{t \, \alpha(t)}
%- \frac{u_i(1)}{t \, \alpha(0)}.
%$$
%
With the inverse-time schedule, Eq. \eqref{eq:alpha_invt},
each term in the sum vanishes,
and if we further define $\alpha(0) \to \infty$,
then
$$
u_i(t+1) = \hat U_i(t),
$$
i.e. the bias potential in this case coincides with
the runtime average of the correction, $\hat u_i(t)$,
at every step, $t$.
%
On the other hand,
with a constant $\alpha(\tau) = a_0$ for $\tau = 0, 1, \dots$,
Eq. \eqref{eq:Uhat_sum} is approximately a plain average\cite{zhou2005}:
$\hat U_i(t) \approx U_i(t)$,
if we can assume an equilibrium condition
with $u_i(t+1) \approx u_i(1)$.
\note{This would be rather risky for metadynamics,
because we know that the certain noise modes
do not easily go away, although the plain average
formula was also recommended for metadynamics.}

We can also cast other updating relations,
such as Eqs. \eqref{eq:wl_update}
and \eqref{eq:vkupdate},
as runtime averages,
but with dynamic weights.
%
The weighted average
%
\begin{align*}
  X_\omega(t)
  =
  \frac{
    \omega(0) \, x(0) + \cdots + \omega(t) \, x(t)
  }
  {
    \Omega(t)
  }
  ,
\end{align*}
%
where
$\Omega(t) \equiv \omega(0) + \cdots + \omega(t)$
is the accumulated weight,
is equivalent to the recurrence relation
%
\begin{align*}
X_\omega(t) = X_\omega(t-1)
  + \frac{\omega(t)}{\Omega(t)}
  \Delta \, x(t)
,
\end{align*}
where
$\Delta x(t) \equiv x(t) - X_\omega(t-1)$
is the proposed change from the previous average
from the data point at step $t$.
%
The relative weight, $\omega(t)/\Omega(t)$,
can be identified as the updating magnitude,
\begin{equation}
  \alpha^*(t) \equiv
  \frac{ \omega(t) } {\Omega(t)}
  =
  1 - \frac{ \Omega(t-1) } {\Omega(t)}
  .
  \label{eq:alpha_from_Omega}
\end{equation}
%
The inverse-time schedule, Eq. \eqref{eq:alpha_invt1},
follows from the case of unweighted average,
$\omega(t) = 1$ (for $t > 0$),
and updating magnitudes above and below this value
correspond to runtime averages that favor future and past data,
respectively.
%
In the case of Eq. \eqref{eq:vkupdate},
we can identify
the bias potential $\tilde v_k(t)$
as $X(t-1)$,
$\tilde f_k(t)$ as $\Delta x(t)$
by Eq. \eqref{eq:uhatt},
and
the mode-$k$ updating magnitude,
$\alpha(t) \, \lambda_k$,
as $\alpha^*(t)$.
%
This mapping shows that
updating magnitude should not exceed $1.0$.
%
One way to satisfy this condition
in converting the results
from the continuous-time setup is
to use Eq. \eqref{eq:alpha_from_Omega}
with
\begin{align*}
  \frac{ \Omega(t) } { \omega(0) }
  = \prod_{\tau=1}^t \frac{1}{1- \alpha^*(\tau)}
\approx \exp\left(\int_0^t \alpha^*(\tau) \, d\tau \right).
\end{align*}

@@

Here, we give some mathematical details
on the homogeneous updating schemes,
%
The updating matrix of the updating schemes, $w_{ij}$,
can be characterized by
a fixed window function, $\mu_0, \mu_1, \dots, \mu_b$,
that shifts with the bin $j$.
%
Here $\mu_l$ denotes the relative updating magnitude
(not adjusted by the boundary correction)
on the $l$th bin from the center.
%
The translational invariance requires $\rho_i$ to be flat,
and the updating matrix, $\mathbf w$,
must be symmetric to satisfy detailed balance,
Eq. \eqref{eq:w_detailedbalance}.
%
The updating matrices share the same set of eigenvectors
(which here are sines and cosines of different frequencies),
and thus are completely determined
by the eigenvalues.
%
Further, the multiplication of $\mathbf w$
can be reduced to a convolution with the window function.
%
We will show below that the eigenvalues of the updating matrix
is related to the window function as a cosine transform.
%
Thus, there is a one-to-one correspondence between
the eigenvalues of $\mathbf w$
and the window function.

@@

%While translationally-invariant
%updating schemes
%naturally suit periodic variables\cite{
%dama2014},
%they can also be extended to non-periodic variables
%by imposing the reflective boundary condition\cite{
%bussi2006}.
%%
%For simplicity, we will assume that the target
%distribution is flat, or $\rho_i = 1/n$, below.

@@

If the width of the Gaussian window, $\sigma_z$,
is much less than $\Delta z$,
then we can extend the limits of the integrals
to infinity, and normalize $\mu(z)$ as
%
\begin{equation}
  \mu(z)
  \approx
  \frac{            1            }
       { \sqrt{ 2 \pi } \sigma_z }
  %
  \exp\left(
        -\frac{   z^2   }
              { 2 \, \sigma_z^2 }
      \right)
  .
\notag
\end{equation}
%
Then, from Eq. \eqref{eq:wband_eigenvalue},
we get Eq. \eqref{eq:lambda_Gaussian},
which shows that all eigenvalues are positive.
%
However,
when discretized on a finite grid,
some eigenvalues can turn negative.
%
We can maintain the convergence of the updating scheme
by redefining the Gaussian scheme
from the same spectrum of eigenvalues, Eq. \eqref{eq:lambda_Gaussian},
and then inversely computing the window function
using Eq. \eqref{eq:mu_from_lambda},
or the updating matrix using Eq. \eqref{eq:w_from_phi}.
%
An example is shown in Fig. \ref{fig:mat}(a).

@@

Consider an ideal FDS simulation under the exact bias potential.
%
By Eq. \eqref{eq:sh_ave},
the fluctuation of the instantaneous histogram
is then given by the random part,
%
\begin{equation}
  f_{*i}(t) = \xi_{*i}(t)
  ,
  \label{eq:f_xi}
\end{equation}
and
%the same holds for the Fourier transform,
$\tilde f_k(t) \approx \tilde \xi_k(t).$
%
If we define the time average,
%
\begin{align*}
  \tilde F_k(T) \equiv \frac 1 T \sum_{t = 1}^T \tilde f_k(t)
  =\mathcal F\left[ \frac{ H_i(T) } { \rho_i } \right]_k
  ,
\end{align*}
%
for the histogram fluctuation of mode $k$,
then the total fluctuation is
%
\begin{align}
  F^2(T)
  &\equiv
  \sum_{k=1}^{n-1} \tilde F_k^2(T)
  =
  \sum_{i=1}^{n} \rho_i \, F_{*i}^2(T)
  =
  \sum_{i=1}^n
  \frac{ [ H_i(T) -\rho_i]^2 }{\rho_i}
  .
  \notag
  %\label{eq:hfluc_def}
\end{align}
%
From Eq. \eqref{eq:f_xi}, we have
%$$\bigl\langle \tilde F_k^2(T) \bigr\rangle
%=
%\sum_{t=-\infty}^\infty
%\frac{
%\bigl\langle
%  \tilde f_k(t) \, \tilde f_k(0)
%\bigr\rangle } { T }
%\approx
%\sum_{t=-\infty}^\infty
%\frac{
%\bigl\langle
%  \tilde \xi_k(t) \, \tilde \xi_k(0)
%\bigr\rangle } { T }
%=
%\frac{ \Gamma_k } { T },$$
%
\begin{align*}
\bigl\langle \tilde F_k^2(T) \bigr\rangle
&=
\frac{1}{T}
\sum_{t=-\infty}^\infty
\bigl\langle
  \tilde f_k(t) \, \tilde f_k(0)
\bigr\rangle
\\
&\approx
\frac{1}{T}
\sum_{t=-\infty}^\infty
\bigl\langle
  \tilde \xi_k(t) \, \tilde \xi_k(0)
\bigr\rangle
=
\frac{ \Gamma_k } { T }
,
\end{align*}
%
and the total histogram fluctuation is
\begin{equation}
  \bigl\langle F^2(T) \bigr\rangle
  =
  \sum_{k=1}^{n-1}
  \bigl\langle \tilde F_k^2(T) \bigr\rangle
  =
  \sum_{k=1}^{n-1}
  \frac{ \Gamma_k } { T }
  =
  \frac{ \Gamma } { T }
  ,
  \label{eq:F2sum}
\end{equation}
where we have used Eq. \eqref{eq:Gammak_sum}
in the last step.
%
We may relate the histogram fluctuation
to the minimal error under the single-bin updating scheme,
Eq. \eqref{eq:Emin_sbin}, as
%we may rephrase the optimality of the inverse-time schedule
%as the inequality,
\begin{equation}
  \Err(T)
  \approx
  \bigl\langle F^2(T) \bigr\rangle
  .
  \notag
  %\label{eq:fluc}
\end{equation}
%and the histogram fluctuation
%serves as a good estimate of the final error.
%

>>>>

%To simplify the ensuing discussion,
Second, we limit ourselves to matrices % $\mathbf w$'s
that satisfy
the detailed balance condition for $\pmb\rho$,
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}
  .
  \notag
  %\label{eq:w_detailedbalance}
\end{equation}
%
This requires the scaled updating matrix,
$\sqrt{ \rho_i/\rho_j } \, w_{ij}$,
to be symmetric,
and the symmetry allows the diagonalization
of $\mathbf w$ with a set of
eigenvectors, $\phi_{ki}$,
%
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij}
  =
  \lambda_k \, \phi_{kj}
  ,
\label{eq:eig_w}
\end{equation}
%
satisfying the orthonormal conditions\cite{vankampen}:
%
\begin{align}
  \sum_{k = 0}^{n - 1}
    \phi_{ki} \, \phi_{kj}
  &=
  \delta_{ij} \, \rho_i,
  \label{eq:eig_orthonormal_cols}
  \\
  \sum_{i = 1}^n
    \frac{ \phi_{ki} \, \phi_{li} }
         { \rho_i }
  &=
  \delta_{kl}
  ,
  \label{eq:eig_orthonormal_rows}
\end{align}
%
where we have started the index of the eigenmodes at $0$
instead of $1$ for later convenience.

@@

\item \label{step:alpha}
Compute the optimal schedule by
numerically integrating Eqs. \eqref{eq:q_opt} and \eqref{eq:mint}.
%
To handle Gaussian updating schemes more effectively,
we have chosen the $\mathcal N+1$ grid points
($\mathcal N=20000$ here)
over $[0, q(T)]$ to be given by
$q_{[i]} = \bigl(q(T) + 1\bigr) - \bigl(q(T)+1\bigr)^{1-i/\mathcal N}$,
where $i = 0, \dots, \mathcal N$.\footnote{For
  a mass distribution, $m(\bar q)$, that is
  inversely proportional to $\bar q$,
  this grid would make the integral,
  or the difference, $t_{[i+1]} - t_{[i]}$,
  roughly a constant.}
%
The curve $q(t)$ is then integrated
using the trapezoidal rule\cite{press3rd},
resulting in a list of values, $\bigl(q_{[i]}, t_{[i]} \bigr)$.
%
We then compute the updating magnitude
from (cf. Appendix \ref{sec:ave_us}):
\begin{equation}
  \alpha_i = \frac{ 1 - e^{-\lambda_1 \, \hat \alpha_i} } { \lambda_1 }
  ,
  \label{eq:alphai}
\end{equation}
where
$\hat \alpha_i \equiv (q_{[i+1]} - q_{[i-1]}) / (t_{[i+1]} - t_{[i-1]})$,
with
$q_{[ -1]} \equiv 0$,
$q_{[\mathcal N+1]} \equiv q(T)$,
$t_{[ -1]} \equiv 0$, and
$t_{[\mathcal N+1]} \equiv T$.
%
The resulting values of $(t_{[i]}, \alpha_{[i]})$
can be linearly interpolated
to get the updating magnitude, $\alpha(t)$, at any simulation time, $t$.

@@

In addition to the above application to
homogeneous updating schemes,
we will consider here a family of bandpass updating schemes
constructed from a sequence of orthogonal polynomials.
%
These updating schemes are more intuitive and easier to implement
as the bias potential in this case is
a low-order polynomial of the collective variable.

@@

For $k > 0$, $R_k(z)$ is a polynomial of degree $k$ with zero expectation.
%
Thus, $R_k\bigl(z(t)\bigr)$ roughly
represents the deviation of the $k$th statistical moment
of the instantaneous distribution
from its expectation.
Since high-order moments are generally less reliable,
this method is best used with a small $K$ for a narrow distribution.
%
However, by constructing a mixture\cite{shirts2017}
or expanded ensemble\cite{swendsen1986,
  *geyer1991, *hukushima1996, *hansmann1997, *sugita1999,
  *earl2005, *zuckerman2011, *rauscher2009,
  neuhaus2006, *neuhaus2007, kim2010,
  marinari1992,
  *lyubartsev1992, li2007,
  park2007, *nguyen2013, *zhang2015st},
we can simultaneously sample multiple target distributions
to cover a wider range, as exemplified below.

@@

In real application, the algorithm is often applied simultaneously
to an array of adjacent intervals
to cover a wider range of the collective variable, $z$.
%
In this way, attempts to move out of the current interval
are not necessarily rejected,
but can be interpreted as transitions to
another internal and accepted with proper probability.

@@

\item
  Change the updating magnitude, $\alpha(t)$,
  which will be used for both intra-umbrella
  parameters, $c_k^{(i)}$,
  and inter-umbrella parameters, $\ln Z^{(i)}$.

@@

After simulation,
the weighted histogram analysis method
(WHAM)\cite{ferrenberg1988, *ferrenberg1989,
*kumar1992, *roux1995, *bartels1997, *souaille2001,
*kobrak2003, *gallicchio2005, *chodera2007,
*fenwick2008, *kim2011, *shirts2008,
zhu2012, newman, frenkel},
can be used to recover the unbiased distribution,
%
\begin{equation}
  p(z) \, \delta z
  =
  \frac{
    \sum_i n^{(i)}(z)
  }
  {
    \sum_i N^{(i)} \, e^{-u^{(j)}(z)} / Z^{(i)}
  }
  ,
  \notag
\end{equation}
%
where $n^{(i)}(z)$
the number of visits to the bin $(z-\delta/2, z+\delta/2)$
in umbrella $i$,
and the $N^{(i)}$ is the total number of visits
to the umbrella $i$.
%
We can use the final converged value of $Z^{(i)}$
without going through WHAM iteration.

@@

\section{\label{sec:equilerr}
Runtime-average interpretation
of the inverse-time schedule
}

We can understand
the optimality of the inverse-time schedule
for the single-bin updating scheme
by a runtime average interpretation.
%
We wish to show that the recurrence updating relation,
Eq. \eqref{eq:wl_update},
renders the bias potential
as a runtime average with a dynamic weight
determined by the updating schedule.
%
For the single-bin updating scheme,
the inverse-time schedule is optimal
because it corresponds to an unweighted average.

\subsection{Runtime average and recurrence relation}

Consider the weighted average
%
\begin{equation}
  X_\omega(t)
  =
  \frac{
    \omega(0) \, x(0) + \cdots + \omega(t) \, x(t)
  }
  {
    \Omega(t)
  }
  ,
  \label{eq:X_ave}
\end{equation}
%
where
$\Omega(t) \equiv \omega(0) + \cdots + \omega(t)$
is the cumulative weight.
%
We can write the average as a recurrence relation
%
\begin{equation}
  X_\omega(t)
  = X_\omega(t-1)
  +
  \alpha(t)
  \, [ x(t) - X_\omega(t-1) ]
  ,
  \label{eq:X_recur}
\end{equation}
%
where the updating magnitude is defined as
%
%The relative weight, $\omega(t)/\Omega(t)$,
%can be identified as the updating magnitude,
\begin{equation}
  \alpha(t) \equiv
  \frac{ \omega(t) } {\Omega(t)}
  =
  1 - \frac{ \Omega(t-1) } {\Omega(t)}
  .
  \label{eq:alpha_from_Omega}
\end{equation}
%
and $x(t) - X_\omega(t-1)$
is the proposed change from the previous average
from the data point at step $t$.

Conversely, for a recurrence relation
in the form of Eq. \eqref{eq:X_recur}
with a given $\alpha(t)$, we can revert it
to an average in the form of Eq. \eqref{eq:X_ave}
with the weight given by
\begin{align}
  \frac{ \omega(t) } { \omega(0) }
  =
  \left(
    \prod_{\tau=1}^{t-1} \frac{1}{1- \alpha(\tau)}
  \right)
  \frac{ \alpha(t) } { 1 - \alpha(t) }
  .
  \label{eq:omega_from_alpha}
\end{align}

%The inverse-time schedule, Eq. \eqref{eq:alpha_invt1},
For example,
the unweighted average, $\omega(t) = \omega(0)/t_0$ (for $t > 0$),
corresponds to the inverse-time updating magnitude,
Eq. \eqref{eq:alpha_invt1}.
%
Updating magnitudes above and below this value
correspond to runtime averages
with growing and diminishing weights, respectively.

\subsection{\label{sec:ave_us}
Application to updating schemes}

To apply the above formalism to updating schemes,
we need a correction formula for the bias potential.
%
We will use a linearized version of Eq. \eqref{eq:vcorr_equil},
$\hat u_i = u_i + \frac{ H_i } { \rho_i } - 1$.
%
%\begin{equation}
%  \hat u_i = u_i + \frac{ H_i } { \rho_i } -1
%  .
%  \label{eq:vcorr_equil1}
%\end{equation}
%
Using this formula to
the instantaneous histogram, $h_i(t)$,
we get an independent estimate
of the bias potential at a single step $t$,
%
\begin{equation}
  \hat u_i(t)
  \approx u_i(t) + \frac{ h_i(t) } { \rho_i } - 1
  = u_i(t) + f_i(t) - 1
  .
  \label{eq:uhatt}
\end{equation}

For the single-bin updating scheme, Eq. \eqref{eq:wl_update},
we have
\begin{align}
  u_i(t+1)
  &=
  u_i(t) + \alpha(t) \, f_i(t)
  \notag \\
  &\approx
  u_i(t) + \alpha(t) \, [\hat u_i(t) - u_i(t)] + \mathrm{const.}
  ,
  \label{eq:u_uhat}
\end{align}
where we have used Eq. \eqref{eq:uhatt}
and ignored a constant shift $\alpha(t)$
in the second step.
%
Comparing this to Eq. \eqref{eq:X_recur},
we find that $u_i(t+1)$
is equivalent to a runtime average $X_\omega(t)$
of $\hat u_i(t)$
with the weight given by Eq. \eqref{eq:omega_from_alpha}.
%
The inverse-time schedule is optimal
because it corresponds to the unweighted average
that uses the data most efficiently.

Similarly, for the multiple-bin updating scheme,
we have, from Eq. \eqref{eq:vkupdate},
${\tilde u}_k(t + 1) = {\tilde u}_k(t) + \alpha(t) \, \lambda_k \, {\tilde f}_k(t)$,
and from Eq. \eqref{eq:u_uhat},
$\hat{\tilde u}_k(t) \approx {\tilde u}_k(t) + {\tilde f}_k(t)$,
for $k > 0$.
This means that
$\tilde u_k(t+1)$ can also be cast
as a runtime average of $\hat{\tilde u}_k(t)$,
with a weight corresponding to the effective updating magnitude,
$\lambda_k \, \alpha(t)$.

The runtime average interpretation imposes
an additional constraint on the updating magnitude.
%
By Eq. \eqref{eq:alpha_from_Omega},
the effective updating magnitude, $\lambda_k \, \alpha(t)$,
of any mode $k > 0$
should never exceed $1.0$,
which is not necessarily satisfied
for the updating schedule
derived from the continuous-time formalism.
%
An expedient fix,
which we used in Eq. \eqref{eq:alphai},
is to derive the effective updating magnitude
from Eq. \eqref{eq:alpha_from_Omega}
with
\begin{align*}
  \ln \frac{ \Omega(t-1) } { \Omega(t) }
  &\approx
  \frac{1}{t-t'} \ln \frac{ \Omega(t') } { \Omega(t) }
  =
  \sum_{\tau=t'+1}^t \frac{ \ln[1- \lambda_k \, \alpha(\tau)] } { t - t' }
  \\
  &\approx
  -\lambda_k \cdot \frac{\int_{t'}^t \alpha(\tau) \, d\tau}{t-t'}
  =
  -\lambda_k \cdot \frac{q(t) - q(t')}{t- t'}
  ,
\end{align*}
where $t$ and $t'$ are two consecutive grid points.
%
Using this formula to mode $k = 1$
with the largest eigenvalue
would guarantee $\lambda_k \, \alpha(t) \ge 1$
for all nontrivial modes.

\subsection{Histogram correction}

We have shown above that
for a general multiple-bin updating scheme,
each mode of bias potential
corresponds to a runtime average
with a dynamic weight,
which is not necessarily even to be most efficient.
%
Below we give a simple correction that
could largely compensate the efficiency loss.
%
We will use capital letters to denote unweighted averages.
%e.g. for variable $y$,
%$Y(t) \equiv \frac{1}{t} \sum_{\tau=1}^t y(\tau)$.
%\begin{equation}
%  Y(t) \equiv \frac{1}{t} \sum_{\tau=1}^t x(\tau)
%  .
%  \notag
%  %\label{eq:timeav}
%\end{equation}
%
Then, by an unweighted average of Eq. \eqref{eq:uhatt}, we get
\begin{align}
  \hat U_i(t)
  &\approx
  U_i (t)
  + F_i(t) - 1
  \approx
  U_i(t)
  + \ln \frac{ H_i(t) } { \rho_i }
  .
  \label{eq:uhatav}
\end{align}
%
This correction serves as
a generalization of Eq. \eqref{eq:vcorr_equil}
for FDS simulations under
a variable bias potential.\footnote{We can also
  derive Eq. \eqref{eq:uhatav}
  from the self-healing umbrella sampling method\cite{marsili2006}.
  We can construct an independent estimator
  of the unbiased distribution from Eq. \eqref{eq:Vi_target} as
  $\hat p_i(t) \approx h_i(t) \, e^{ u_i(t) }$,
  assuming a proper shift of $u_i(t)$.
  Then after time averaging, we get
  $\rho_i e^{\hat U_i(t)} = \hat P_i(t) \approx H_i \, e^{ U_i(t) }$.}

@@

Here, we discuss some details
on the homogeneous updating schemes.
%
We can characterize
the updating matrix, $w_{ij}$,
of a homogeneous updating scheme
by a rigid window function
that shifts with the bin $j$ of visit
(ignoring the boundary adjustment).
%
The window function is defined by
the relative updating magnitude, $\mu_l$,
on the $l$th bin from the central bin.
%
%The translational invariance requires $\rho_i$ to be flat,
%and the updating matrix, $\mathbf w$,
%must be symmetric to satisfy detailed balance,
%Eq. \eqref{eq:w_detailedbalance}.
%
By the translational invariance,
the updating matrices of homogeneous updating schemes
share the same set of eigenvectors,
which are sine and cosine functions.
%
The eigenvalues of the updating matrix
are related to the window function as a cosine transform.

@@

@@

If the width of the Gaussian window, $\sigma_z$,
is much less than $\Delta z$,
then we can extend the limits of the integrals
to infinity, and normalize $\mu(z)$ as
%
\begin{equation}
  \mu(z)
  \approx
  \frac{            1            }
       { \sqrt{ 2 \pi } \sigma_z }
  %
  \exp\left(
        -\frac{   z^2   }
              { 2 \, \sigma_z^2 }
      \right)
  .
\notag
\end{equation}
%
Then, from Eq. \eqref{eq:wband_eigenvalue},
we get Eq. \eqref{eq:lambda_Gaussian},
which shows that all eigenvalues are positive.
%
However,
when discretized on a finite grid,
some eigenvalues can turn negative.
%
To maintain the positivity of eigenvalues,
we can redefine the Gaussian scheme
from the same spectrum of eigenvalues, Eq. \eqref{eq:lambda_Gaussian},
and then inversely computing the window function
using Eq. \eqref{eq:mu_from_lambda},
or the updating matrix using Eq. \eqref{eq:w_from_phi}.
%
An example is shown in Fig. \ref{fig:mat}(a).

@@

Consider an ideal FDS simulation under the exact bias potential.
%
By Eq. \eqref{eq:sh_ave},
the fluctuation of the instantaneous histogram
is then given by the random part,
%
\begin{equation}
  f_{*i}(t) = \xi_{*i}(t)
  ,
  \label{eq:f_xi}
\end{equation}
and
%the same holds for the Fourier transform,
$\tilde f_k(t) = \tilde \xi_k(t).$
%
If we define the time average,
%
\begin{align*}
  \tilde F_k(T) \equiv \frac 1 T \sum_{t = 1}^T \tilde f_k(t)
  =\mathcal F\left[ \frac{ H_i(T) } { \rho_i } \right]_k
  ,
\end{align*}
%
for the histogram fluctuation of mode $k$,
then the total fluctuation is
%
\begin{align}
  F^2(T)
  &\equiv
  \sum_{k=1}^{n-1} \tilde F_k^2(T)
  =
  \sum_{i=1}^{n} \rho_i \, F_{*i}^2(T)
  =
  \sum_{i=1}^n
  \frac{ [ H_i(T) -\rho_i]^2 }{\rho_i}
  .
  \notag
  %\label{eq:hfluc_def}
\end{align}
%
From Eq. \eqref{eq:f_xi}, we have
%$$\bigl\langle \tilde F_k^2(T) \bigr\rangle
%=
%\sum_{t=-\infty}^\infty
%\frac{
%\bigl\langle
%  \tilde f_k(t) \, \tilde f_k(0)
%\bigr\rangle } { T }
%\approx
%\sum_{t=-\infty}^\infty
%\frac{
%\bigl\langle
%  \tilde \xi_k(t) \, \tilde \xi_k(0)
%\bigr\rangle } { T }
%=
%\frac{ \Gamma_k } { T },$$
%
\begin{align*}
\bigl\langle \tilde F_k^2(T) \bigr\rangle
&=
\frac{1}{T}
\sum_{t=-\infty}^\infty
\bigl\langle
  \tilde f_k(t) \, \tilde f_k(0)
\bigr\rangle
\\
&\approx
\frac{1}{T}
\sum_{t=-\infty}^\infty
\bigl\langle
  \tilde \xi_k(t) \, \tilde \xi_k(0)
\bigr\rangle
=
\frac{ \Gamma_k } { T }
,
\end{align*}
%
and the total histogram fluctuation is
\begin{equation}
  \bigl\langle F^2(T) \bigr\rangle
  =
  \sum_{k=1}^{n-1}
  \bigl\langle \tilde F_k^2(T) \bigr\rangle
  =
  \sum_{k=1}^{n-1}
  \frac{ \Gamma_k } { T }
  =
  \frac{ \Gamma } { T }
  ,
  \label{eq:F2sum}
\end{equation}
where we have used Eq. \eqref{eq:Gammak_sum}
in the last step.
%
We may relate the histogram fluctuation
to the minimal error under the single-bin updating scheme,
Eq. \eqref{eq:Emin_sbin}, as
%we may rephrase the optimality of the inverse-time schedule
%as the inequality,
\begin{equation}
  \Err(T)
  \approx
  \bigl\langle F^2(T) \bigr\rangle
  .
  \notag
  %\label{eq:fluc}
\end{equation}
%and the histogram fluctuation
%serves as a good estimate of the final error.
%
